---
name: categorical-ex
---


---

## [Ex.]{.ex} Card (1995)

Education is not randomly assigned across the population, it is a choice. [_"Depending on how these choices are made, measured earnings differences between workers with different levels of schooling may over-state or under-state the true return to education."_]{.fragment}

<br>

. . .

[Card (1995)]{.hii} uses geographic information to causally identify the impact of education earnings by comparing young men who grew up near higher education institutions to those who did not:

<br>

. . .

Findings suggests the greatest earnings increases are among poor men[, suggesting that the presence of a local college lowers the costs/raises the perceived benefits of education.]{.fragment}

---

## [Ex.]{.ex} Card (1995) Abstract

:::{.vertical-center}
> Although schooling and earnings are highly correlated, social scientists have argued for decades over the causal effect of education. [This paper explores the use of college proximity as an exogenous determinant of schooling.]{.hii} An examination... reveals that [men who grew up in local labor markets with a nearby college have significantly higher education and significantly higher earnings than other men]{.hii}. The education and earnings gains are concentrated among men with poorly- educated parents -- men who would otherwise stop schooling at relatively low levels.
:::

---

## [Ex.]{.ex} Card (1995) Data {data-visibility="uncounted"}

::: {.tiny}
| Variable   | Description                                   |
|------------|-----------------------------------------------|
| id         | Person identifier                             |
| nearc4     | =1 if near 4 yr college, 1966                 |
| educ       | Years of schooling, 1976                      |
| age        | Age in years                                  |
| fatheduc   | Father's schooling                            |
| motheduc   | Mother's schooling                            |
| weight     | NLS sampling weight, 1976                     |
| black      | =1 if black                                   |
| south      | =1 if in south, 1976                          |
| wage       | Hourly wage in cents, 1976                    |
| IQ         | IQ score                                      |
| libcrd14   | =1 if lib. card in home at 14                 |
:::

---

<!-- ## [Ex.]{.ex} Card (1995) -->

::: {.small}

Regress wages on an indicator for proximity to a four year institution

```{r}
#| echo: true
lm(wage ~ nearc4, data = wage_data) %>% tidy()
```

:::: {.columns .tiny}

::: {.column width="40%"}
[Q~1~:]{.note} What is the reference category? 

[Q~2~:]{.note} Interpret the coefficients.

<br>
<br>
<br>
<br>

[Q~3~:]{.note} Suppose instead we run^[where `farc4` is an indicator for living far from 4-year institutions] 

```r
lm(wage ~ farc4, data = wage_data
```

How does interpretation change?

:::

::: {.column width="60%"}
[_Men who did not grow up near a 4 year institution_]{.fragment}

---

[$\beta_0$: _On average, men who did not grow up near a 4-year institution earn 516 cents per hour._]{.fragment}

[$\widehat{\beta}_{\text{nearc4}}$: _On average, men who grew up near a 4 year institution earn 89.2 cents more per hour than men who did not._]{.fragment}

---

[$\beta_0$: _On average, men who grow up near a 4-year institution earn $\widehat{\beta}_0$ cents per hour._]{.fragment}

[$\widehat{\beta}_{\text{farc4}}$: _On average, men who grew up far away from a 4 year institution earn $\widehat{\beta}_{farc4}$ cents less per hour than men who did._]{.fragment}
:::

::::

:::

---

::: {.tiny}

Regress wages on an indicator for proximity to a four year institution and race

```{r}
#| echo: true
lm(wage ~ nearc4 + black, data = wage_data) %>% tidy()
```

:::: {.columns}

::: {.column width="30%"}
[Q~1~:]{.note} Reference category? 

[Q~2~:]{.note} Interpretation?


:::

::: {.column width="70%"}
[_Non-black men who did not grow up near a 4-year institution_]{.fragment .tiny}

---

[$\beta_0$: _On average, non-black men who grew up far from a 4-year institution earn 562 cents per hour._]{.fragment .tiny}

[$\widehat{\beta}_{\text{nearc4}}$: _On average, men who grew up near a 4-year institution earn 78.1 cents more per hour than men who did not._]{.fragment .tiny}

[$\widehat{\beta}_{\text{black}}$: _On average, black men earn 162 cents less per hour than non-black men, holding proximity to a 4-year institution constant._]{.fragment .tiny}

:::

::::

:::

---

::: {.tiny}

Regress wages on an indicator for proximity to a four year institution and race

```{r}
#| echo: true
lm(wage ~ nearc4 + nonblack, data = wage_data) %>% tidy()
```

:::: {.columns}

::: {.column width="30%"}
[Q~1~:]{.note} Reference category? 

[Q~2~:]{.note} Interpretation?

:::

::: {.column width="70%"}
[_Black men who did not grow up near a 4-year institution_]{.fragment .tiny}

---

[$\beta_0$: _On average, black men who did not grow up near a 4-year institution earn 400 cents per hour._]{.fragment .tiny}

[$\widehat{\beta}_{\text{nearc4}}$: _On average, men who grew up near a 4-year institution earn 78.1 cents more per hour than men who did not._]{.fragment .tiny}

[$\widehat{\beta}_{\text{black}}$: _On average, nonblack men earn 162 cents more per hour than non-black men, holding proximity to a 4-year institution constant._]{.fragment .tiny}

:::

::::

:::

---

::: {.middle .tiny}
_Notice:_

- The coefficient on `nearc4` is the same in both models
- The interpretation of the coefficient on `black` changes---magnitude is the same


::: {.fragment}

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
lm(wage ~ nearc4 + black, data = wage_data) %>% 
  tidy()
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
lm(wage ~ nearc4 + nonblack, data = wage_data) %>% 
  tidy()
```
:::

::::

:::
:::


---

## [Ex.]{.ex} Card (1995)

```{r}
#| echo: true

lm(wage ~ educ + nearc4 + south + black, data = wage_data) %>% tidy()
```

[Q~1~:]{.note} What is the reference category?

[Q~2~:]{.note} Interpret the coefficients.

---

We considered a model where schooling has the same effect for everyone ([F]{.hp} and [M]{.hi})

```{r}
#| include: false
#| cache: true

# Set seed
set.seed(12345)
# Sample size
n <- 1e3
# Parameters
beta0 <- 20; beta1 <- 0.5; beta2 <- 10; beta3 <- 3
# Dataset
int_df <- tibble(
  male = sample(x = c(F, T), size = n, replace = T),
  school = runif(n, 3, 9) - 3 * male,
  pay = beta0 + beta1 * school + beta2 * male + rnorm(n, sd = 7) + beta3 * male * school
)
reg_noint <- lm(pay ~ school + male, int_df)
reg_int <- lm(pay ~ school + male + school:male, int_df)
```

```{r}
#| dev: svg
#| echo: false
#| fig.height: 5.5

ggplot(data = int_df, aes(x = school, y = pay)) +
geom_point(aes(color = male, shape = male), size = 2.5) +
geom_abline(
  intercept = reg_noint$coefficients[1] + reg_noint$coefficients[3],
  slope = reg_noint$coefficients[2],
  color = hi, size = 1, alpha = 0.8
) +
geom_abline(
  intercept = reg_noint$coefficients[1],
  slope = reg_noint$coefficients[2],
  color = hp, size = 1, alpha = 0.8
) +
xlab("Schooling") +
ylab("Pay") +
coord_cartesian(xlim = c(0, 10), ylim = c(0, 80)) +
mytheme_s +
theme(
  text = element_text(color = hi, family = "Fira Sans"),
  axis.title = element_text(size = 18),
  plot.margin = structure(c(0, 0, 0.1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
) +
scale_color_manual("", values = c(hp, hi), labels = c("Female", "Male")) +
scale_shape_manual("", values = c(16, 1), labels = c("Female", "Male"))
```

---

We will consider models that allow effects to differ by another variable (*e.g.,* by gender: ([F]{.hp} and [M]{.hi})):

```{r}
#| dev: svg
#| echo: false
#| fig.height: 5.5
ggplot(data = int_df, aes(x = school, y = pay)) +
geom_point(aes(color = male, shape = male), size = 2.5) +
geom_abline(
  intercept = reg_noint$coefficients[1] + reg_noint$coefficients[3],
  slope = reg_noint$coefficients[2],
  color = hi, size = 0.75, alpha = 0.2
) +
geom_abline(
  intercept = reg_noint$coefficients[1],
  slope = reg_noint$coefficients[2],
  color = hp, size = 0.75, alpha = 0.2
) +
geom_abline(
  intercept = reg_int$coefficients[1] + reg_int$coefficients[3],
  slope = reg_int$coefficients[2] + reg_int$coefficients[4],
  color = hi, size = 1, alpha = 0.8
) +
geom_abline(
  intercept = reg_int$coefficients[1],
  slope = reg_int$coefficients[2],
  color = hp, size = 1, alpha = 0.8
) +
xlab("Schooling") +
ylab("Pay") +
coord_cartesian(xlim = c(0, 10), ylim = c(0, 80)) +
mytheme_s +
theme(
  text = element_text(color = hi, family = "Fira Sans"),
  axis.title = element_text(size = 18),
  plot.margin = structure(c(0, 0, 0.1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
) +
scale_color_manual("", values = c(hp, hi), labels = c("Female", "Male")) +
scale_shape_manual("", values = c(16, 1), labels = c("Female", "Male"))
```
