---
name: ovb
---

---

## Omitted variable bias

> Bias that occurs in statistical models when a relevant variable is not included in the model.


. . .

<br>

[Consequence:]{.hi-red} Leads to the incorrect estimation of the relationships between variables, which may affect the reliability of the model's predictions and inferences.

<br>

. . .

[Solution:]{.hii} [_"Control"_]{.note} for the omitted variable(s).



---

## {data-visibility="uncounted"}

Class funding ([U]{.hi}) [confounds]{.note} our estimates of smaller class sizes ([X]{.hi}) on test scores ([Y]{.hi}). 
<br>
<br>


<!-- DAG SLIDE -->

```{r}
#| label: dag-ex-ovb-022-again
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6
#| fig-align: center

# Plot the full DAG
ggplot(
  data = dag_dt,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)
```

_Any unobserved variable_ that connects a [backdoor path]{.note} between class size ([X]{.hi}) and test scores ([Y]{.hi}) will [bias]{.hi} our [point estimate]{.note} of $\beta_1$.

---

## {data-visibility="uncounted"}

Class funding ([U]{.hi}) [confounds]{.note} our estimates of smaller class sizes ([X]{.hi}) on test scores ([Y]{.hi}). Including data on school funding ([U]{.hi}) in a multiple linear regression allows us to close this backdoor path.

```{r}
#| label: dag-ex-ovb-033
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6
#| fig-align: center

# Add indicators for paths
dag_dt[, `:=`(
  path1 = (name == "X" & to == "Y") | (name == "Y"),
  path2 = (name == "X" & to == "U") | (name == "U" & to == "Y") | (name == "Y")
)]

# Add alpha level
dag_dt[, `:=`(
  alpha = ifelse((name == "X" & to == "U") | (name == "U" & to == "X"), 0.3, 1)
)]

# Continue with the rest of your code...


# Plot the full DAG
ggplot(
  data = dag_dt,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb, alpha = alpha),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)
```

. . .

With all backdoor paths closed, [point estimates]{.note} of $\beta_1$ will no longer be biased and will return the population parameter of interest

---

::: {.middle}
_In a little more detail, we can derive the bias mathematically._
:::

---

## Omitted Variable Bias

Imagine we have a population model of the form:

$$
Y_i = \beta_0 + \beta_1 X_i + \beta_2 Z_i + u_i
$$

where $Z_i$ is a relevant variable that is omitted from the model.

and suppose we estimate the following model:

$$
Y_i = \hat{\beta}_0 + \hat{\beta}_1 X_i + v_i
$$

where $v_i$ is the new error term that absorbs the effect of $Z_i$


---

## Omitted Variable Bias

To derive the bias of $\hat{\beta}_1$, we need to understand the relationship between $Z_i$ and $X_i$. Assume that:

$$
Z_i = \gamma_0 + \gamma_1 X_i + \varepsilon_i
$$

where $\varepsilon_i$ is the part of $Z_i$ that is uncorrelated with $X_i$

. . .

<br>

If we substitute $Z_i$ into the population model, we get:

$$
\begin{align*}
Y_i &= \beta_0 + \beta_1 X_i + \beta_2 \left( \gamma_0 + \gamma_1 X_i + \varepsilon_i \right) + u_i \\
    &= \beta_0 + \beta_2 \gamma_0 + \left( \beta_1 + \beta_2 \gamma_1 \right) X_i + \beta_2 \varepsilon_i + u_i
\end{align*}
$$

---

## Omitted Variable Bias

We can rewrite this expression:

$$
\begin{align*}
Y_i &= \beta_0 + \beta_1 X_i + \beta_2 \left( \gamma_0 + \gamma_1 X_i + \varepsilon_i \right) + u_i \\
    &= \beta_0 + \beta_2 \gamma_0 + \left( \beta_1 + \beta_2 \gamma_1 \right) X_i + \beta_2 \varepsilon_i + u_i
\end{align*}
$$

as:

$$
Y_i = \widehat{\beta}_0 + \widehat{\beta}_1 X_i + v_i
$$

where:

- $\widehat{\beta}_0 = \beta_0 + \beta_2 \gamma_0$
- $\widehat{\beta}_1 = \beta_1 + \beta_2 \gamma_1$
- $v_i = \beta_2 \varepsilon_i + u_i$

Thus, we can see how $Z_i$ will bias our estimate of $\beta_1$

---

## Omitted Variable Bias

Recall that we define the bias of an estimator as:

$$
\mathop{\text{Bias}}_\theta \left( W \right) = \mathop{\boldsymbol{E}}\left[ W \right] - \theta
$$

. . .

The bias of the estimator $\hat{\beta}_1$ is given by:

$$
\begin{align*}
\mathop{\text{Bias}}_{\beta_1} \left( \hat{\beta}_1 \right) &= \mathop{\boldsymbol{E}}\left[ \hat{\beta}_1 \right] - \beta_1 \\
&= \mathop{\boldsymbol{E}}\left[ \beta_1 + \beta_2 \gamma_1 \right] - \beta_1 \\
&= \beta_2 \gamma_1
\end{align*}
$$

---

## Omitted Variable Bias

Finally, we can write the bias of $\hat{\beta}_1$ in terms of the correlation between $X_i$ and $Z_i$:

$$
\gamma_1 = \frac{\text{Cov}\left( X_i, Z_i \right)}{\text{Var}\left( X_i \right)}
$$

. . .

Therefore, we can write the bias of $\hat{\beta}_1$ as:

$$
\mathop{\text{Bias}}_{\beta_1} \left( \hat{\beta}_1 \right) = \beta_2 \frac{\text{Cov}\left( X_i, Z_i \right)}{\text{Var}\left( X_i \right)}
$$

Sure, let's convert the explanation to use the expectation \(\mathop{\boldsymbol{E}}\) instead of the probability limit \(\mathop{\text{plim}}\).

---

## Signing the Bias

Sometimes we're stuck with omitted variable bias.

$$
\mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( X_i,\, Z_i \right)}{\mathop{\text{Var}} \left( X_i \right)}
$$

When this happens, we can often at least know the direction of the bias.

---

## Signing the Bias {data-visiblity="uncounted"}

Begin with

$$
\mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( X_i,\, Z_i \right)}{\mathop{\text{Var}} \left( X_i \right)}
$$

We know $\color{#8FBCBB}{\mathop{\text{Var}} \left( X_i \right) > 0}$. Suppose $\color{#81A1C1}{\beta_2 > 0}$ and $\color{#EBCB8B}{\mathop{\text{Cov}} \left( X_i,\,Z_i \right) > 0}$. Then

. . .

$$
\begin{align}
 \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \color{#81A1C1}{(+)} \dfrac{\color{#EBCB8B}{(+)}}{\color{#8FBCBB}{(+)}} \implies \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] > \beta_1
\end{align}
$$
∴ In this case, OLS is **biased upward** (estimates are too large).

. . .

$$
\begin{matrix}
 \enspace & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)> 0} & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)< 0} \\
 \color{#81A1C1}{\beta_2 > 0} & \text{Upward} &  \\
 \color{#81A1C1}{\beta_2 < 0} &  &
\end{matrix}
$$

---

## Signing the Bias {data-visiblity="uncounted"}

Begin with

$$
\mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( X_i,\, Z_i \right)}{\mathop{\text{Var}} \left( X_i \right)}
$$

We know $\color{#8FBCBB}{\mathop{\text{Var}} \left( X_i \right) > 0}$. Suppose $\color{#81A1C1}{\beta_2 < 0}$ and $\color{#EBCB8B}{\mathop{\text{Cov}} \left( X_i,\,Z_i \right) > 0}$. Then

. . .

$$
\begin{align}
 \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \color{#81A1C1}{(-)} \dfrac{\color{#EBCB8B}{(+)}}{\color{#8FBCBB}{(+)}} \implies \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] < \beta_1
\end{align}
$$
∴ In this case, OLS is **biased downward** (estimates are too small).

$$
\begin{matrix}
 \enspace & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)> 0} & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)< 0} \\
 \color{#81A1C1}{\beta_2 > 0} & \text{Upward} &  \\
 \color{#81A1C1}{\beta_2 < 0} & \text{Downward} &
\end{matrix}
$$

---

## Signing the Bias {data-visiblity="uncounted"}

Begin with

$$
\mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( X_i,\, Z_i \right)}{\mathop{\text{Var}} \left( X_i \right)}
$$

We know $\color{#8FBCBB}{\mathop{\text{Var}} \left( X_i \right) > 0}$. Suppose $\color{#81A1C1}{\beta_2 > 0}$ and $\color{#EBCB8B}{\mathop{\text{Cov}} \left( X_i,\,Z_i \right) < 0}$. Then

$$
\begin{align}
 \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \color{#81A1C1}{(+)} \dfrac{\color{#EBCB8B}{(-)}}{\color{#8FBCBB}{(+)}} \implies \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] < \beta_1
\end{align}
$$
∴ In this case, OLS is **biased downward** (estimates are too small).

$$
\begin{matrix}
 \enspace & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)> 0} & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)< 0} \\
 \color{#81A1C1}{\beta_2 > 0} & \text{Upward} & \text{Downward} \\
 \color{#81A1C1}{\beta_2 < 0} & \text{Downward} &
\end{matrix}
$$

---

## Signing the Bias {data-visiblity="uncounted"}

Begin with

$$
\mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( X_i,\, Z_i \right)}{\mathop{\text{Var}} \left( X_i \right)}
$$

We know $\color{#8FBCBB}{\mathop{\text{Var}} \left( X_i \right) > 0}$. Suppose $\color{#81A1C1}{\beta_2 < 0}$ and $\color{#EBCB8B}{\mathop{\text{Cov}} \left( X_i,\,Z_i \right) < 0}$. Then

$$
\begin{align}
 \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] = \beta_1 + \color{#81A1C1}{(-)} \dfrac{\color{#EBCB8B}{(-)}}{\color{#8FBCBB}{(+)}} \implies \mathop{\boldsymbol{E}} \left[ \hat{\beta}_1 \right] > \beta_1
\end{align}
$$
∴ In this case, OLS is **biased upward** (estimates are too large).

$$
\begin{matrix}
 \enspace & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)> 0} & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)< 0} \\
 \color{#81A1C1}{\beta_2 > 0} & \text{Upward} & \text{Downward} \\
 \color{#81A1C1}{\beta_2 < 0} & \text{Downward} & \text{Upward}
\end{matrix}
$$

---

## Signing the Bias

Thus, in cases where we have a sense of

1. the sign of $\mathop{\text{Cov}} \left( X_i,\,Z_i \right)$

2. the sign of $\beta_2$

we know in which direction bias pushes our estimates.

**Direction of Bias**

$$
\begin{matrix}
 \enspace & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)> 0} & \color{#EBCB8B}{\text{Cov}(X_i,\,Z_i)< 0} \\
 \color{#81A1C1}{\beta_2 > 0} & \text{Upward} & \text{Downward} \\
 \color{#81A1C1}{\beta_2 < 0} & \text{Downward} & \text{Upward}
\end{matrix}
$$