---
title: "**Lab 01 - Intro to R and RStudio**"
author: "Dante Yasui"
date: "2024/04/03"
output: pdf_document
--- 

# Welcome to Lab!

Goals for this week:

- Install R and RStudio 
- Learn about R basics and vectors 
- Install the tidyverse package
- Learn how to use and complete Koans  
- **Complete and submit `K01_vector.R`**
  * not due until next Friday
  
---

# Setting up our R environment

## Step 1. Download and install R 

follow instructions from: <https://cran.r-project.org/>

- choose latest release for your OS
- Mac users: 
  * know if your device has Apple silicon (M1-3 macs) or Intel cpu
  * also install xquartz: <https://www.xquartz.org/>

## Step 2. Download and install RStudio

Go here: <https://posit.co/download/rstudio-desktop/>

- click on the link under **2. Install RStudio**
- follow the installer instructions

Open RStudio 

You should see the default panes layout

- find the `Console` in the bottom left pane
- you can use this to run R code

---

#### R as a fancy calculator

```{r}
1 + 2
```
```{r}
sqrt(64)
```
#### Everything is an object to R

Objects have a name and value(s)

```{r}
# we can define variables with either `<-` or `=`
a <- 9
b = 3
a / b
```

This is useful to for all sorts of reasons

  * labeling things with memorable names
  * having functions call on values which might change
  * using shorter names for long values

e.g.,
```{r}
pi <- 3.141593
```

#### There are different *types* of objects
```{r}
name     = 'dante'
age      = 25
hometown = "Davis, CA"
```

```{r}
typeof(name)
```
```{r}
typeof(age)
```

```{r}
typeof(hometown)
```

#### Vectors are lists

To `combine` things into a vector, use the `c()` syntax:
```{r}
c(1,2,3)
```

Vectors are also objects:
```{r}
stooges <- c("larry", "moe", "curly")
stooges
```


#### Functions do things with objects
```{r}
secret_message = "hello world!"
print(secret_message)
```

You can also define your own functions:
```{r}
my_addition <- function(a, b) {
  return (a + b - 1)
}
```
  
R is really good at doing things with vector objects

```{r}
long_vector <- 1:100
long_vector
```
```{r}
long_vector ^ 2
```

---

## Step 3. Install packages

User-defined functions come in **Packages**

#### Install the tidyverse

Use the R function `install.packages()` to install the tidyverse from CRAN
```{r, eval=FALSE}
install.packages("tidyverse", dependencies = TRUE)
```

Don't forget to always load your package before using:
```{r}
library(tidyverse)
```

#### Get beginner-friendly help from `qelp`

Created by Colleen O'Briant: <https://github.com/cobriant/qelp>

```{r, eval=F}
# we need these to install qelp:
install.packages("Rcpp", dependencies = TRUE)
install.packages("devtools", dependencies = TRUE)
```

answer `'yes'` when prompted in the console

Now you can install qelp:
```{r, message=FALSE}
library(devtools)
install_github("cobriant/qelp")
```

Test that it worked:
```{r}
?qelp::install.packages
```

You can also use the default `R` documentation:
```{r}
?install.packages
```

See the difference?

## Step 4. Download the Koans

Go to the github page: <https://github.com/ajdickinson/tidyverse_koans>

Click the `<> Code` button, then `Download ZIP`.

Unzip the files wherever you want them on your personal machine. 
This will be where you will keep all of your lab work for this class.

Double-click the file called `tidyverse_koans-master.Rproj`.

This should open a new `Project` in RStudio with all of the R files you need.

Open the `R` folder from the files you downloaded
by clicking it in the `Files` tab in RStudio.
Now click on the first lab assignment, `K01_vector.R`,
and it will open in your `source` tab where you can view and edit it.

---

# Complete Koan 1 - Vectors

You can use the hotkey `Ctrl/Cmd Shift C` to comment out the lines of code
in between each set of question markers (e.g., from `#1@` to `@#1`).

Fill in the blanks by following the instructions. 

#### Test your answers

Use the shortcut `Shift Ctrl/Cmd T` to check whether your code is correct.

#### Compile and submit to Canvas

Once your code has passed all of the tests,
you will compile the R script as an html formatted output
which you will upload to Canvas.

In the RStudio menu bar, go to *Tools*, then *Modify Keyboard Shortcuts*.
Set the shortcut for *Compile Notebook* to be `Shift Ctrl/Cmd K`
(or whatever shortcut you will remember).

You can also use the menu option in `File` for `Knit Document`.

Once RStudio is done compiling,
you will have a file called `K01_vector.html` 
in the same folder as the original `.R` file.

Upload this to Canvas to get credit for this lab.
