---
name: assumptionsA4-A6
---

## Classical Assumptions of OLS {data-visibility="uncounted"}

[A1.]{.note} [Linearity:]{.hi} The population relationship is [_linear in parameters_]{.note} with an additive error term.

[A2.]{.note} [Sample Variation:]{.hi} There is variation in $X$.

[A3.]{.note} [Exogeniety:]{.hi} The $X$ variable is [exogenous]{.note} 

. . .

<br>
<br>

::: {.align-center}
The following 2 assumptions are not required for unbiasedness...
:::

::: {.align-center}
[But the are important for an efficient estimator]{.fragment}
:::

---

##

:::{.vertical-center}
[_Variance matters, too_]{.note}
:::

---

## Why variance matters

Unbiasedness tells us that OLS gets it right, _on average_. [But we can't tell whether our sample is "typical."]{.fragment}

<br>

. . .

[Variance]{.hi} tells us how far OLS can deviate from the population mean.

- How tight is OLS centered on its expected value?
- This determines the [efficiency]{.hp} of our estimator.

---

## Why variance matters {data-visibility="uncounted"}

Unbiasedness tells us that OLS gets it right, _on average_. But we can't tell whether our sample is "typical."

<br>

The smaller the variance, the closer OLS gets, _on average_, to the true population parameters _on any sample_.

- Given two unbiased estimators, we want the one with smaller variance.
- If [two more assumptions]{.note} are satisfied, we are using the [most efficient]{.hp} linear estimator.

---

## Classical Assumptions of OLS {data-visibility="uncounted"}

[A1.]{.note} [Linearity:]{.hi} The population relationship is [_linear in parameters_]{.note} with an additive error term.

[A2.]{.note} [Sample Variation:]{.hi} There is variation in $X$.

[A3.]{.note} [Exogeniety:]{.hi} The $X$ variable is [exogenous]{.note} 

. . .

[A4.]{.note} [Homoskedasticity:]{.hi} The error term has the same variance for each value of the independent variable 

---

## Homoskedasticity ([A4.]{.note})

> The error term has the same variance for each value of the independent variable:

$$
\mathop{Var}(u|X) = \sigma^2.
$$

[Example:]{.hi}

```{r}
#| echo: false
#| fig.height: 3
#| fig.align: center

set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4)
), aes(x = x, y = e)) +
geom_point(color = hi, size = 2.75, alpha = 0.5) +
labs(x = "Independent Variable", y = "Error Term") +
mytheme_s +
theme(
  axis.text.x = element_text(size = 12),
  axis.title.x = element_text(size = 12),
  axis.text.y = element_blank()
)

```

---

## Homoskedasticity ([A4.]{.note})

> The error term has the same variance for each value of the independent variable:

$$
\mathop{Var}(u|X) = \sigma^2.
$$

[Violation:]{.hi} [Heteroskedasticity]{.note}

```{r}
#| echo: false
#| fig.height: 3
#| fig.align: center

set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4 + 1.5 * x)
), aes(x = x, y = e)) +
geom_point(color = hi, size = 2.75, alpha = 0.5) +
labs(x = "Independent Variable", y = "Error Term") +
mytheme_s +
theme(
  axis.text.x = element_text(size = 12),
  axis.title.x = element_text(size = 12),
  axis.text.y = element_blank()
)

```

---

## Homoskedasticity ([A4.]{.note})

> The error term has the same variance for each value of the independent variable:

$$
\mathop{Var}(u|X) = \sigma^2.
$$

[Violation:]{.hi} [Heteroskedasticity]{.note}

```{r}
#| echo: false
#| fig.height: 3
#| fig.align: center

set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2)
), aes(x = x, y = e)) +
geom_point(color = hi, size = 2.75, alpha = 0.5) +
labs(x = "Independent Variable", y = "Error Term") +
mytheme_s +
theme(
  axis.text.x = element_text(size = 12),
  axis.title.x = element_text(size = 12),
  axis.text.y = element_blank()
)

```

---

## Heteroskedasticity [Ex.]{.ex}

Suppose we study the following relationship:

$$
\text{Luxury Expenditure}_i = \beta_1 + \beta_2 \text{Income}_i + u_i
$$

<br>

As income increases, variation in luxury expenditures [increase]{.hii} 

- Variance of $u_i$ is likely larger for higher-income households
- Plot of the residuals against the household income would likely reveal a funnel-shaped pattern

--- 

## {data-visibility="uncounted"}

[Common test for heteroskedasticity...]{.fragment} [Plot the residuals across $X$]{.fragment}

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, 0, 10),
  e = rnorm(1e3, 0, sd = 4 + 1.5 * x)
), aes(x = x, y = e)) +
geom_point(color = hi, size = 2.75, alpha = 0.5) +
labs(x = "Income", y = "Residuals") +
scale_x_continuous(breaks = seq(0,10,2)) +
mytheme_s +
theme(
  axis.text.x = element_text(size = 12),
  axis.title.x = element_text(size = 12),
  axis.text.y = element_blank()
)

```

---

## {data-visibility="uncounted"}

::: {.vertical-center}
[There is more to be said about homo/heteroskedasticity in EC 421]{.note}
:::

---

## Classical Assumptions of OLS {data-visibility="uncounted"}

[A1.]{.note} [Linearity:]{.hi} The population relationship is [_linear in parameters_]{.note} with an additive error term.

[A2.]{.note} [Sample Variation:]{.hi} There is variation in $X$.

[A3.]{.note} [Exogeniety:]{.hi} The $X$ variable is [exogenous]{.note} 

[A4.]{.note} [Homoskedasticity:]{.hi} The error term has the same variance for each value of the independent variable 

. . .

[A5.]{.note} [Non-autocorrelation:]{.hi} The values of error terms have independent distributions 

---

## Non-Autocorrelation 

> The values of error terms have independent distributions^[[Notes:]{.note} $\forall i = \text{for all} \: i$, $\text{s.t.} = \text{such that}$, $i \neq j \: \text{means} \: i \: \text{is not equal to} \: j$]

$$
E[u_i u_j]=0, \forall i \text{ s.t. } i \neq j
$$

. . .

Or...

$$
\begin{align*}
\mathop{\text{Cov}}(u_i, u_j) &= E[(u_i - \mu_u)(u_j - \mu_u)]\\
                              &= E[u_i u_j] = E[u_i] E[u_j]  = 0, \text{where } i \neq j
\end{align*}
$$

---

## Non-Autocorrelation 

> The values of error terms have independent distributions 

$$
E[u_i u_j]=0, \forall i \text{ s.t. } i \neq j
$$

- Implies no systematic association between pairs of individual $u_i$
- Almost always some unobserved correlation across individuals^[(e.g. common correlation in unobservables among individuals within a given US state)]
- Referred to as [clustering]{.hp} problem.
- An easy solution exists where we can adjust our standard errors

## {data-visibility="uncounted"}

::: {.vertical-center}
Let's take a moment to talk about the [variance]{.note} of the [OLS]{.hi} [estimator]{.note}
:::

---

[Proof of the population variance]{.note}

By definition

$$
Var(\hat{\beta}_2) = \mathbb{E}\left[\left(\hat{\beta}_2 - E(\hat{\beta}_2)\right)^2\right] = \mathbb{E}\left[\left(\hat{\beta}_2 - \beta_2\right)^2\right]
$$

. . .

[Step 1:]{.note} Derive an expression for $\hat{\beta}_2 - \beta_2$:

$$
\hat{\beta}_2 - \beta_2 = \frac{\sum(y_i - \bar{y})(x_i - \bar{x})}{\sum(x_i - \bar{x})^2} - \beta_2
$$

. . .

Substitute for $Y_i$:

$$
\hat{\beta}_2 - \beta_2 = \frac{\sum((\beta_1 + \beta_2x_i + u_i) - \bar{y})(x_i - \bar{x})}{\sum(x_i - \bar{x})^2} - \beta_2
$$

---

[Proof of the population variance]{.note}

Substitute $\bar{y} = \beta_1 + \beta_2\bar{x}$:

$$
\hat{\beta}_2 - \beta_2 = \frac{\sum(u_i(x_i - \bar{x}))}{\sum(x_i - \bar{x})^2}
$$

. . .

[Step 1:]{.note} Take variance $\hat{\beta}_2 - \beta_2$:

$$
Var(\hat{\beta}_2) = \mathbb{E}\left[\left(\frac{\sum(u_i(x_i - \bar{x}))}{\sum(x_i - \bar{x})^2}\right)^2\right]
$$

---

Square the numerator:

$$
Var(\hat{\beta}_2) = \mathbb{E}\left[\frac{\sum(u_i(x_i - \bar{x}))\sum(\epsilon_j(x_j - \bar{x}))}{\left(\sum(x_i - \bar{x})^2\right)^2}\right]
$$

. . .

Using [A5]{.note}, we can simplify:

$$
Var(\hat{\beta}_2) = \frac{1}{\left(\sum(x_i - \bar{x})^2\right)^2}\sum \mathbb{E}[u_i^2(x_i - \bar{x})^2]
$$

---

Using [A4]{.note}: $Var(u_i) = \sigma^2$.

$$
Var(\hat{\beta}_2) = \frac{\sigma^2}{\left(\sum(x_i - \bar{x})^2\right)^2}\sum (x_i - \bar{x})^2
$$

. . .

<br>

Thus we arrive at the variance of the OLS slope parameter:

$$
Var(\hat{\beta}_2) = \frac{\sigma^2}{\sum(x_i - \bar{x})^2}
$$


---

## {data-visibility="uncounted"}

::: {.vertical-center}
If [A4.]{.note} and [A5.]{.note} are satisfied, along with [A1.]{.note}, [A2.]{.note}, and [A3.]{.note}, then we are using the [most efficient]{.hi} linear estimator.
:::

---

## Classical Assumptions of OLS {data-visibility="uncounted"}

[A1.]{.note} [Linearity:]{.hi} The population relationship is [_linear in parameters_]{.note} with an additive error term.

[A2.]{.note} [Sample Variation:]{.hi} There is variation in $X$.

[A3.]{.note} [Exogeniety:]{.hi} The $X$ variable is [exogenous]{.note} 

[A4.]{.note} [Homoskedasticity:]{.hi} The error term has the same variance for each value of the independent variable 

[A5.]{.note} [Non-autocorrelation:]{.hi} The values of error terms have independent distributions 

. . .

[A6.]{.note} [Normality:]{.hi} The population error term in normally distributed with mean zero and variance $\sigma^2$

---

## Normality ([A6.]{.note})

> The population error term in normally distributed with mean zero and variance $\sigma^2$

Or,

$$
u \sim N(0,\sigma^2)
$$

Where $\sim$ stands for [distributed by]{.note} and $N$ stands for [normal distribution]{.note}

. . .

<br>

However, [A6.]{.note} is note required for efficientcy nor unbiasedness

---

